Fabric-Elasticity Relationships of Tibial Trabecular Bone are Similar in Osteogenesis Imperfecta and Healthy Individuals

### 2021-07-02 09_Article/Simon_2021_Bone_Draft_V02.tex
Finish article corrections

	In the present study, the morphology of distal tibiae from 50 adults with OI were compared to 120 healthy controls using second generation HR-pQCT. Six cubic regions of interest (ROIs) were selected per individual in a common anatomical region. A first matching between OI and healthy control group was performed by selecting similar individuals to obtain identical mean and median age and gender distribution. It allowed to perform a morphometric analysis and compare the outcome with literature. Then, stiffness tensors of ROIs were computed using hFE and multiple linear regressions were performed with the Zysset-Curnier orthotropic model. The regressions allowed to compare the two groups using five parameters. An initial fit was performed on both the OI group and the healthy control group using all extracted ROIs. Then, data was filtered according to a fixed threshold for a defined coefficient of variation (CV) assessing the ROI heterogeneity and additionnal linear regressions were performed on these filtered data sets. These full and filtered data were in turn compared with previous results from \si{\micro}CT reconstructions obtained in other anatomical locations. Finally, the ROIs of both groups were matched according to their BV/TV and fabric anisotropy (DA). Linear regressions were performed using these matched data to detect statistical differences between the two groups.

@@

	In conclusion, despite the reduced regression parameters found for HR-pQCT images, the fabric-elasticity relationships between OI and healthy individuals are similar when the trabecular bone ROIs are sufficiently homogeneous to perform the mechanical analysis. Since highly heterogenous ROIs coincide with very low BV/TV, we expect them to play a minor role in hFE analysis of distal bone sections or parts.
\end{abstract}

@@

\subsection{Participants}
The healthy group included a total of 120 patients from a previous reproducibility study performed at the University Department of osteoporosis in Bern \cite{Schenk2020}. The group was composed of 64 females and 56 males aged between 20 and 92 years old with a mean age of 32 $\pm$ 15 years. These subjects had not taken any medication known to affect bone metabolism nor presented with any prior osteoporosis fracture. The second group was scanned as part of the ASTEROID study at different locations in Canada, namely at the Shriners Hospital for Children-Canada. The study coordination was done by the McGill University in Montreal. This group was composed of 35 females and 15 males with confirmed diagnosis of OI Type I, III or IV. There were 35 patients diagnosed with OI type I, 2 with type III, and 13 with type IV. The participants of the OI group were aged between 19 and 69 years old with a mean age of 44 $\pm$ 14 years.

\subsection{HR-pQCT}
HR-pQCT scans (XtremeCTII, SCANCO Medical, Brütisellen,
Switzerland) were performed at the distal tibia on all patients from both groups. Participants in the healthy group were scanned using an in-house protocol as described in \cite{Schenk2020}. Namely, the reference line was positioned at the proximal margin of the dense structure formed by the tibia plafond and three stacks were scanned proximal to this reference line, see Figure \ref{01_Healthy}. On the other hand, participants in the OI group were scanned using the manufacturer's standard protocol i.e. the reference line is placed at the subchondral endplate of the ankle joint and one stack was scanned at 22.5 mm proximal to the reference line (standard clinical section) \cite{Whittier2020}, see Figure \ref{01_OI}. The scanned region according to the two different protocol are shown side by side in Figure \ref{01_ClinicalSections}. For both the healthy and the OI groups, each stack consisted of 168 slices and a voxel resolution of 61 \si{\micro}m in the three principal directions. This led to a thickness of roughly 10.2 mm for each stack. Standardized scanning settings were used (voltage of 60 kVp, 900 μA, 100 ms integration time) for the healthy group as well as for the OI group. For the healthy group, motion artefacts of first, middle and last slice (i.e slices number 1, 252, and 504) were graded on a scale of 1 (no motion artefacts) to 5 (extreme motion artefacts), as proposed by \citeauthor{Pialat2012} \cite{Pialat2012}. The final grade of each scan was defined as the highest slice grade. For the OI group, as the scan consisted of one stack, only one grade was attributed using the same scale as for the control group. Scans were then processed independently of their quality grading. A summary of the scans grading is shown in Figure \ref{01_MotionArtefacts}.

@@

In general, six cubic ROIs were selected at random position in each scan in a defined area. Each ROI had to contain trabecular bone, but no cortical bone. For the healthy group, the ROIs were selected in the more proximal stack uniquely (see Figure \ref{01_Healthy} stack number 3) to be at the same anatomical location as for the OI group. Then for both the healthy group and the OI group, the stack was divided into two halves and the ROIs were selected to have the centres of three ROIs in the proximal half and three in the distal half. For one individual diagnosed with OI type III it was not possible to extract any ROI as there was no enough trabecular bone. This led to 720 healthy ROIs from 120 individuals and 294 OI ROIs from 49 individuals.\\

The ROI was defined as a cube of 5.3 mm side length. This size was in agreement with the work of \citeauthor{Panyasantisuk2015}\cite{Panyasantisuk2015} and \citeauthor{Gross2013}\cite{Gross2013}, who performed similar analysis with femur \si{\micro}CT scans. It was determined by \citeauthor{Zysset1998}\cite{Zysset1998} and \citeauthor{Daszkiewicz2017}\cite{Daszkiewicz2017} to allow one to have a relative homogeneity of trabecular tissue within the ROI leading to accurate \si{\micro}FE results with a minimal computational cost.\\

After ROI cleaning, i.e. deletion of unconnected region of bone material, the morphological analysis of ROIs was performed using medtool (v4.5; Dr. Pahr Ingenieurs e.U., Pfaffstätten, Austria). The morphological parameters analyzed were: BV/TV, structural model index (SMI), trabecular number (Tb.N.), trabecular thickness (Tb.Th.), trabecular separation (Tb.Sp.), and the standard deviation of the trabecular spacing (Tb.Sp.SD). Moreover, ROI tissue bone mineral density (tBMD) and fabric was evaluated. The fabric tensor $\mathbf{M}$ was computed using mean intercept length (MIL) method \cite{Moreno2014}. It is a positive-definite second-order tensor built as shown in Equation \ref{Eq201} below:

@@

A \si{\micro}FE mechanical analysis was performed using \textsc{ABAQUS 6.14}. In brief, each voxel of the cleaned ROI was converted to a fully integrated linear brick elements (C3D8) using a direct voxel conversion approach. Then, a stiffness $E$ of 10,000 MPa and a Poisson's ratio $\nu$ of 0.3 were assigned. The homogenization process consisted of 6 independent simulations of different load cases, 3 uni-axial and 3 simple shear cases, using kinematic uniform boundary conditions (KUBCs) \cite{Panyasantisuk2015}. Unlike periodicity-compatible mixed uniform boundary conditions (PMUBCs), KUBCs do not require one to rotate the ROI into fabric coordinate system. Such rotation would potentially decrease image quality. The homogenization process allowed to calculate the components of the stiffness tensor and to calibrate the parameters of the Zysset-Curnier fabric-elasticity model \cite{Zysset1995}. This model builds the fourth order stiffness tensor $\mathbb{S}$ using the BV/TV or $\rho$, fabric tensor $\mathbf{M}$, three elasticity parameters $\lambda_0$, $\lambda_0$', and $\mu_0$, and two exponents, $k$ and $l$, as shown in Equation \ref{Eq202}.\\

@@

Where $\mathbb{S}'$ and $\mathbb{S}$ are the transformed and the original stiffness tensor respectively and $Q$ is the orthogonal matrix that maps the original coordinate system into the new one (fabric). The Zysset-Curnier model is built with the assumption of orthotropy and homogeneity. However, the trabecular structure is not perfectly homogeneous. In order to assess the ROI heterogeneity, a coefficient of variation (CV) is computed according to \citeauthor{Panyasantisuk2015}\cite{Panyasantisuk2015}: the ROI is divided into eight identical sub-cubes and BV/TV is computed for each of them. The CV is defined as the ratio between the standard deviation of these BV/TV and the mean value (Equation \ref{Eq203}).

@@

\subsection{Statistics}
The analyzed morphological parameters (BV/TV, Tb.N., Tb.Th., Tb.Sp., and Tb.Sp.SD, SMI, DA, and CV) were compared between the healthy and the OI groups. As the groups do not have similar distributions of age and sex, a matching was performed by selecting similar individuals leading to identical mean and median age as well as identical gender distribution. For each parameter, the median value between the six ROIs from the same individual was computed. The median was preferred over the mean because it is less influenced by outliers. Normality of the distribution was assessed with QQ plot and Shapiro-Wilk test. CV had to be log-transformed to meet normal distribution assumption. Then, equal variances was assessed using Bartlett test or Brown-Forsythe test according to the normality distribution of the data. According to the normality and equal variances assumptions, t-test, Mann-Whitney test or a non-parametric permutation test was performed. The general significance level was set to 95\% for all tests. Confidence intervals were computed for t-tested variables to quantify the difference in both groups means. As Mann-Whitney tests are performed on the median, only the corresponding p-value is presented. Finally, non-parametric permutation tests are less powerful but give an empirical 95\% exclusion range and a p-value. If the difference in means belong to this exclusion range, it can be stated that group means are different with 95\% certainty.

@@

Where $S_{rc}$ is the $r$th row and $c$th column of the non-zero element of the orthotropic stiffness tensor $\mathbb{S}$ in Mandel notation \cite{MANDEL1965}, $X$ is a $12n$x$p$ design matrix containing the the BV/TV and fabric info of the $n$ ROIs and $\beta$ is a $p$x1 vector of fixed effects containing model parameters. $Z$ is a $12n$x$f$ design matrix which contains data with individual dependence and $\delta$ is a $f$x1 vector composed of random factors. Finally, $\epsilon$ is a $12n$x1 vector containing the regression residuals. As $k$ and $l$ are exponents, the regression was performed on the log space.\\

The linear regression was performed on both group (heal\-thy and OI) separately. To improve the fit quality, the data sets were filtered. The aim here was to filter out ROIs violating the assumption of homogeneity. Therefore, analogously to the work of \citeauthor{Panyasantisuk2015}\cite{Panyasantisuk2015}, a fixed threshold for the CV was used. To simplify comparison, we used the same value 0.263 as exclusion criterion \cite{Panyasantisuk2015}. Then, the relation between BV/TV and CV was assessed using Spearman's correlation coefficient. To compare the stiffness constants ($\lambda_0$, $\lambda_0'$, and $\mu_0$) between the groups, regression must be performed on identical value ranges. Therefore, a matching was performed for BV/TV and DA to find corresponding control ROIs for each OI in the filtered groups. Best correspondences were kept and duplicates were dropped. Finally, as the regression was performed in the log space, slight differences in the exponents ($k$ and $l$) would lead to important variation of the stiffness constants ($\lambda_0$, $\lambda_0'$, and $\mu_0$) so it was necessary to use identical exponents  for both groups, weighting identically BV/TV and DA between regressions. The exponents were determined by grouping healthy and OI for regression. Then a modified system was used to perform the fit on separated groups, see Appendix \ref{A1}, Equation \ref{EqA11}.\\

Another modification of the model was to add a regressor for the group variable (healthy or OI), i.e. add a column to the design matrix $X$ and a row to the parameter vector $\beta$. This modified model is compared to the original by analysis of covariance (ANCOVA) using the fixed-effects only to determine the statistical significance of the group. Implementation of this modification was performed according to \cite{Fox2016}. A similar mixed-effect model was used to analyze the relation between tBMD and BV/TV and the significance of the group, see Appendix \ref{A1} Equation \ref{EqA14}. The model used the BV/TV and the group (healthy or OI) as fixed variables and the individual as random variable. Moreover, to test the hypothesis of no interaction between the BV/TV and the group, i.e. the group has no significant influence on the tBMD versus BV/TV slope, the model was modified to add the interaction regressor (BV/TV x Group). The detailed linear systems for each model discussed here are available in Appendix \ref{A1} and a summary of the data sets used for the different methods is shown in Table \ref{Table1}.\\

The regression was performed using the \textsc{statsmodels} package from \textsc{Python 3.6}. Regression quality for the tBMD analysis was assessed using the Pearson correlation coefficient ($R^2$) and the standard error of the estimate (SE). Regression on Zysset-Curnier model was assessed using the adjusted Pearson correlation coefficient squared ($R^2_{adj}$) and relative error between the orthotropic observed and the predicted tensor using norm of fourth-order tensors (NE), see Equation \ref{Eq205} and \ref{Eq206}.

@@

\subsection{Morphological Analysis}
The results of the morphological analysis are summarized in Table \ref{Table2}. The individual matching for age and sex allowed us to have similar group distributions with 17 females and 11 males in each group. The mean age of matched healthy individuals is 41y $\pm$ 14y and 41y $\pm$ 15y for the mat\-ched OI individuals. BV/TV of healthy individuals is higher than BV/TV of OI group with a difference 95\% CI of [0.016, 0.101] and p-value <0.01. Similarly, trabecular number is higher in the matched healthy group compared to the matched OI group with a difference 95\% CI of [0.099, 0.285] and a corresponding p-value <0.001. The trabecular thickness does not show significant differences between groups. On the other hand, permutation test performed for trabecular separation showed that trabecular separation is higher in mat\-ched OI group compared to healthy individuals with a p-value of 0.01 and an exclusion range of ($-\infty$ ,-0.384] $\cup$ [0.421,$\infty$). Trabecular separation SD is higher in OI individuals compared to matched healthy individuals with a p-value of 0.02 and an exclusion range of ($-\infty$ ,-0.232] $\cup$ [0.251,$\infty$). SMI as well as the degree of anisotropy are higher for matched OI than for healthy people with p-values <0.001 and of 0.02, respectively. Finally, the log transformation of the coefficient of variation gives the stronger difference in means with a p value <0.0001 and a 95\% CI of [−0.757, −0.333] where CV is higher in matched OI individuals compared to matched healthy individuals.\\

@@

\subsection{Linear Regression with Original Data Sets}
Figure \ref{02_GeneralRegression} shows the results of the linear regression analysis of each group separately, between the values of the observed stiffness tensors from \si{\micro}FE simulations and the predicted values using the Zysset-Curnier model\cite{Zysset1995} and the parameters obtained after performing the regression with linear mixed-effect model. The fitted line is represented by the dashed line, indicating the theoretical perfect correlation. For the healthy group (Figure \ref{02_Healthy}) the fit is performed on 720 ROIs leading to 8640 data points. The $R^2_{adj}$ is slightly above 0.95 and the NE is of 18\% $\pm$ 10\%. The regression analysis for the OI group (Figure \ref{02_OI}) performed on 294 ROIs led to 3528 data points, an $R^2_{adj}$ close to 0.85 and a NE of 62\% $\pm$ 233\%. It can be noticed that, as the values of the observed stiffness tensor decreases, data points tend to be further apart from the diagonal (dashed line). Moreover, data the points from the stiffness tensors with lowest values are exclusively above the diagonal. The range of stiffness tensors of the OI group is wider compared to the one of the healthy group and ROIs with lower BV/TV present lower stiffness values. The values of these ROIs stiffness tensors components tends to be overestimated by the fit.\\

\begin{figure}[h!]
	\centering
	\begin{subfigure}[b]{0.5\textwidth}
		\centering
		\includegraphics[width=\textwidth]
		{Pictures/02_GR_Healthy_LMM}
		\caption{Healthy group}
		\label{02_Healthy}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.5\textwidth}
		\centering
		\includegraphics[width=\textwidth]
		{Pictures/02_GR_OI_LMM}
		\caption{OI group}
		\label{02_OI}
	\end{subfigure}
	\caption{Regression results using the fixed effects of the linear mixed-effect model on original data sets. $\lambda_{ii}$ stands for the diagonal terms of normal components of $\mathbb{S}$ in Mandel notation\cite{MANDEL1965}, $\lambda_{ij}$ for the off-diagonal terms of normal components, and $\mu_{ij}$ for the shear components. The dashed line represents the fitted line.}
	\label{02_GeneralRegression}
\end{figure}

\subsection{Filtering}
The CV in relation to BV/TV is shown in Figure \ref{02_CV_BVTV}. The OI data reached higher CV values and lower BV/TV values compared to healthy data. Generally, the CV tends to increase with decreasing BV/TV. The Spearman coefficient is shown above the plot as value [95\% CI]. Its value is negative and strictly different from zero. Finally, the CV threshold value used to filter the data is represented by the dashed line. It can be observed that a relatively important part of OI data will be filtered out. On the other hand, relatively few healthy data gets removed. 3D representation of extreme ROIs in terms of CV and BV/TV are shown in Appendix \ref{A2}.\\

\begin{figure}[h!]
	\centering
	\includegraphics[width=\linewidth]
	{Pictures/03_CV_BVTV}
	\caption{Coefficient of variation in relation to BV/TV. Spearman correlation coefficient $\rho$ assess monotonic relation between two variable}
	\label{02_CV_BVTV}
\end{figure}

The regression results of the filtered data are presented in Figure \ref{04_FilteredRegression}. After filtering, the healthy group was reduced to 119 individuals and 603 ROIs resulting in 7236 data points, an $R^2_{adj}$ close to 0.95 and a NE of 16\% $\pm$ 8\% (Figure \ref{04_Healthy}). In the OI group, more individuals were filtered leading to 38 people, 115 ROIs, and 1380 data points (Figure \ref{04_OI}). This resulted in an $R^2_{adj}$ close to 0.95 and a NE of 17\% $\pm$ 8\%.\\

>>>>

	In the present study, the morphology of distal tibiae from 50 adults with OI were compared to 120 healthy controls using second generation HR-pQCT. Six cubic regions of interest (ROIs) were selected per individual in a common anatomical region. A first matching between OI and healthy control group was performed by selecting similar individuals to obtain identical mean and median age and gender distribution. It allowed to perform a morphometric analysis and compare the outcome with literature. Then, stiffness tensors of ROIs were computed using hFE and multiple linear regressions were performed with the Zysset-Curnier orthotropic model. The linear regressions allowed to compare the two groups using five parameters. An initial fit was performed on both the OI group and the healthy control group using all extracted ROIs. Then, data was filtered according to a fixed threshold for a defined coefficient of variation (CV) assessing the ROI heterogeneity and additionnal linear regressions were performed on these filtered data sets. These full and filtered data were in turn compared with previous results from \si{\micro}CT reconstructions obtained in other anatomical locations. Finally, the ROIs of both groups were matched according to their BV/TV and fabric anisotropy (DA). Linear regressions were performed using these matched data to detect statistical differences between the two groups.

@@

	In conclusion, despite the reduced linear regression parameters found for HR-pQCT images, the fabric-elasticity relationships between OI and healthy individuals are similar when the trabecular bone ROIs are sufficiently homogeneous to perform the mechanical analysis. Since highly heterogenous ROIs coincide with very low BV/TV, we expect them to play a minor role in hFE analysis of distal bone sections or parts.
\end{abstract}

@@

\subsection{Participants}
The healthy group included a total of 120 patients from a previous reproducibility study performed at the University Department of osteoporosis in Bern \cite{Schenk2020}. The group was composed of 64 females and 56 males aged between 20 and 92 years old with a mean age of 32 $\pm$ 15 years. These subjects had not taken any medication known to affect bone metabolism nor presented with any prior osteoporosis fracture. The second group was scanned as part of the ASTEROID study (NCT number: NCT03118570) at different locations, namely at the Shriners Hospital for Children-Canada. The study coordination was done by the McGill University in Montreal. This group was composed of 35 females and 15 males with confirmed diagnosis of OI Type I, III or IV. There were 35 patients diagnosed with OI type I, 2 with type III, and 13 with type IV. The participants of the OI group were aged between 19 and 69 years old with a mean age of 44 $\pm$ 14 years.

\subsection{HR-pQCT}
HR-pQCT scans (XtremeCTII, SCANCO Medical, Brütisellen,
Switzerland) were performed at the distal tibia on all patients from both groups. Participants in the healthy group were scanned using an in-house protocol as described in \cite{Schenk2020}. Namely, the reference line was positioned at the proximal margin of the dense structure formed by the tibia plafond and three stacks were scanned proximal to this reference line, see Figure \ref{01_Healthy}. On the other hand, participants in the OI group were scanned using the manufacturer's standard protocol i.e. the reference line is placed at the subchondral endplate of the ankle joint and one stack was scanned at 22.5 mm proximal to the reference line (standard clinical section) \cite{Whittier2020}, see Figure \ref{01_OI}. The scanned region according to the two different protocol are shown side by side in Figure \ref{01_ClinicalSections}. For both the healthy and the OI groups, each stack consisted of 168 slices and an isotropic voxel resolution of 61 \si{\micro}m. This led to a thickness of roughly 10.2 mm for each stack. Standardized scanning settings were used (voltage of 60 kVp, 900 μA, 100 ms integration time) for the healthy group as well as for the OI group. For the healthy group, motion artefacts of first, middle and last slice (i.e slices number 1, 252, and 504) were graded on a scale proposed by \citeauthor{Pialat2012} \cite{Pialat2012} of 1 (no motion artefacts) to 5 (extreme motion artefacts). The final grade of each scan was defined as the highest slice grade. For the OI group, as the scan consisted of one stack, only one grade was attributed using the same scale as for the control group. Scans were then processed independently of their quality grading. A summary of the scans grading is shown in Figure \ref{01_MotionArtefacts}.

@@

In general, six cubic ROIs were selected at random position in each scan in a predefined area. Each ROI had to contain trabecular bone, but no cortical bone. For the healthy group, the ROIs were selected in the most proximal stack uniquely (see Figure \ref{01_Healthy} stack number 3) to be at rougthly the same anatomical location as for the OI group. Then for both the healthy group and the OI group, the stack was divided into two halves and the ROIs were selected to have the centers of three ROIs in the proximal half and three in the distal half, see Figure \ref{01_ROISelection}. For one individual diagnosed with OI type III it was not possible to extract any ROI as there was no enough trabecular bone. This led to 720 healthy ROIs from 120 individuals and 294 OI ROIs from 49 individuals.\\

\begin{figure}[h!]
	\centering
	\includegraphics[width=\linewidth]
	{Pictures/01_ROISelection_Example}
	\caption{Middle slices example of ROIs selection for HR-pQCT reconstruction of a healthy tibia. Cortical bone is in white and trabecular bone in gray. The stacks are separated by red lines and blue squares represent the ROIs. The stack used for analysis is higlighted in green and the dashed red line represents the middle of the stack.}
	\label{01_ROISelection}
\end{figure}

The ROI was defined as a cube of 5.3 mm side length. This size agreed with the work of \citeauthor{Panyasantisuk2015}\cite{Panyasantisuk2015} and \citeauthor{Gross2013}\cite{Gross2013}, who performed similar analysis with femur \si{\micro}CT scans. It was determined by \citeauthor{Zysset1998}\cite{Zysset1998} and \citeauthor{Daszkiewicz2017}\cite{Daszkiewicz2017} to allow a relative homogeneity of trabecular tissue within the ROI leading to accurate \si{\micro}FE results and a minimal computational cost.\\

After ROI cleaning, i.e. deletion of bone material regions unconnected to the main structure, the morphological analysis of ROIs was performed using medtool (v4.5; Dr. Pahr Ingenieurs e.U., Pfaffstätten, Austria). The morphological parameters analyzed were: BV/TV, structural model index (SMI), trabecular number (Tb.N.), trabecular thickness (Tb.Th.), trabecular separation (Tb.Sp.), and the standard deviation of the trabecular spacing (Tb.Sp.SD). Moreover, ROI tissue mineral density (TMD) and fabric was evaluated. The fabric tensor $\mathbf{M}$ was computed using mean intercept length (MIL) method \cite{Moreno2014}. It is a positive-definite second-order tensor computed as shown below.

@@

A \si{\micro}FE analysis was performed using \textsc{ABAQUS 6.14}. In brief, each voxel of the cleaned ROI was converted to a fully integrated linear brick elements (C3D8) using a direct voxel conversion approach. Then, a stiffness $E$ of 10,000 MPa and a Poisson's ratio $\nu$ of 0.3 were assigned to each element. The homogenization process consisted of 6 independent simulations of different load cases, 3 uni-axial and 3 simple shear cases, using kinematic uniform boundary conditions (KUBCs) \cite{Panyasantisuk2015}. Unlike periodicity-compatible mixed uniform boundary conditions (PMUBCs), KUBCs do not require to rotate the ROI into the respective fabric coordinate system. Such rotation would potentially decrease image quality. The homogenization process allowed to calculate the components of the stiffness tensor and to calibrate the parameters of the Zysset-Curnier fabric-elasticity model \cite{Zysset1995}. This model compute the fourth order stiffness tensor $\mathbb{S}$ using the BV/TV or $\rho$, fabric tensor $\mathbf{M}$, three elasticity parameters $\lambda_0$, $\lambda_0$', and $\mu_0$, and two exponents, $k$ and $l$, as shown in Equation \ref{Eq202}.\\

@@

Where $\mathbb{S}'$ and $\mathbb{S}$ are the transformed and the original stiffness tensor respectively and $Q$ is the orthogonal matrix that maps the original coordinate system into the new one (fabric). The Zysset-Curnier model is built with the assumption of orthotropy and homogeneity. However, the trabecular structure is not perfectly homogeneous. To assess the ROI heterogeneity, a coefficient of variation (CV) is computed according to \citeauthor{Panyasantisuk2015}\cite{Panyasantisuk2015}: the ROI is divided into eight identical sub-cubes and BV/TV is computed for each of them. The CV is defined as the ratio between the standard deviation of these BV/TV and the mean value (Equation \ref{Eq203}).

@@

\subsection{Morpholigical Analysis}
The analyzed morphological parameters (BV/TV, Tb.N., Tb.Th., Tb.Sp., and Tb.Sp.SD, SMI, DA, and CV) were compared between the healthy and the OI group.

\subsubsection{Participants Matching}
As the groups do not have similar distributions of age and sex, a matching was performed by selecting similar individuals leading to identical mean and median age as well as identical gender distribution.

\subsubsection{Statistical Tests}
For each parameter, the median value between the six ROIs from the same individual was computed. The median was preferred over the mean because it is less influenced by outliers. Normality of the distribution was assessed with QQ plot and Shapiro-Wilk test. CV had to be log-transformed to meet normal distribution assumption. Then, equal variances was assessed using Bartlett test or Brown-Forsythe test according to the normality distribution of the data. According to the normality and equal variances assumptions, t-test, Mann-Whitney test or a non-parametric permutation test was performed.

\subsubsection{Statistical Significance}
The general significance level was set to 95\% for all tests. Confidence intervals were computed for t-tested variables to quantify the difference in both groups means. As Mann-Whitney tests are performed on the median, only the corresponding p-value is presented. Finally, non-parametric permutation tests are less powerful but give an empirical 95\% exclusion range and a p-value. If the difference in means belong to this exclusion range, it can be stated that group means are different with 95\% certainty.

@@

Where $S_{rc}$ is the $r$th row and $c$th column of the non-zero element of the orthotropic stiffness tensor $\mathbb{S}$ in Mandel notation \cite{MANDEL1965}, $X$ is a $12n$x$p$ design matrix containing the the BV/TV and fabric info of the $n$ ROIs and $\beta$ is a $p$x1 vector of fixed effects containing model parameters. $Z$ is a $12n$x$f$ design matrix which contains data with individual dependence and $\delta$ is a $f$x1 vector composed of random factors. Finally, $\epsilon$ is a $12n$x1 vector containing the linear regression residuals. As $k$ and $l$ are exponents, the linear regression was performed on the log space.\\

\subsubsection{Data Filtering}
The linear regression was performed on both group (heal\-thy and OI) separately. To improve the fit quality, the data sets were filtered. The aim here was to filter out ROIs violating the assumption of homogeneity. Therefore, analogously to the work of \citeauthor{Panyasantisuk2015}\cite{Panyasantisuk2015}, a fixed threshold for the CV was used. To simplify comparison, we used the same value 0.263 as exclusion criterion \cite{Panyasantisuk2015}. Then, the relation between BV/TV and CV was assessed using Spearman's correlation coefficient.

\subsubsection{ROIs Matching}
To compare the stiffness constants ($\lambda_0$, $\lambda_0'$, and $\mu_0$) between the groups, linear regression must be performed on identical value ranges. Therefore, a matching was performed for BV/TV and DA to find corresponding control ROIs for each OI in the filtered groups. Best correspondences were kept and duplicates were dropped. Finally, as the linear regression was performed in the log space, slight differences in the exponents ($k$ and $l$) would lead to important variation of the stiffness constants ($\lambda_0$, $\lambda_0'$, and $\mu_0$) so it was necessary to use identical exponents  for both groups, weighting identically BV/TV and DA between linear regressions. The exponents were determined by grouping healthy and OI for linear regression. Then a modified system was used to perform the fit on separated groups, see Appendix \ref{A1}, Equation \ref{EqA11}.\\

\subsubsection{Model Modifications}
Another modification of the model was to add a regressor for the group variable (healthy or OI), i.e. add a column to the design matrix $X$ and a row to the parameter vector $\beta$. This modified model is compared to the original by analysis of covariance (ANCOVA) using the fixed-effects only to determine the statistical significance of the group. Implementation of this modification was performed according to \cite{Fox2016}. A similar mixed-effect model was used to analyze the relation between TMD and BV/TV and the significance of the group, see Appendix \ref{A1} Equation \ref{EqA14}. The model used the BV/TV and the group (healthy or OI) as fixed variables and the individual as random variable. Moreover, to test the hypothesis of no interaction between the BV/TV and the group, i.e. the group has no significant influence on the TMD versus BV/TV slope, the model was modified to add the interaction regressor (BV/TV x Group). The detailed linear systems for each model discussed here are available in Appendix \ref{A1} and a summary of the data sets used for the different methods is shown in Table \ref{Table1}.\\

The linear regression was performed using the \textsc{statsmodels} package from \textsc{Python 3.6}. Linear regression quality for the TMD analysis was assessed using the Pearson correlation coefficient ($R^2$) and the standard error of the estimate (SE). Linear regression on Zysset-Curnier model was assessed using the adjusted Pearson correlation coefficient squared ($R^2_{adj}$) and relative error between the orthotropic observed and the predicted tensor using norm of fourth-order tensors (NE), see Equation \ref{Eq205} and \ref{Eq206}.

@@

\subsection{Morphological Analysis}
The results of the morphological analysis are summarized in Table \ref{Table2}. The individual matching for age and sex allowed to have similar group distributions with 17 females and 11 males in each group. The mean age was 41y $\pm$ 14y and 41y $\pm$ 15y for the mat\-ched healthy and OI individuals, respectively. BV/TV of healthy individuals was higher than BV/TV of OI group with a difference 95\% CI of [0.016, 0.101] and p-value <0.01. Similarly, trabecular number was higher in the matched healthy group compared to the matched OI group with a difference 95\% CI of [0.099, 0.285] and a corresponding p-value <0.001. The trabecular thickness does not show significant differences between groups. On the other hand, permutation test performed for trabecular separation showed that it is higher in mat\-ched OI group compared to healthy individuals with a p-value of 0.01 and an exclusion range of ($-\infty$, -0.384] $\cup$ [0.421, $\infty$). Trabecular separation SD is higher in OI individuals compared to matched healthy individuals with a p-value of 0.02 and an exclusion range of ($-\infty$, -0.232] $\cup$ [0.251, $\infty$). SMI as well as the degree of anisotropy are higher for matched OI than for healthy people with p-values <0.001 and of 0.02, respectively. Finally, the log transformation of the coefficient of variation gives the stronger difference in means with a p value <0.0001 and a 95\% CI of [−0.757, −0.333] where CV is higher in matched OI individuals compared to matched healthy individuals.\\

@@

\subsection{Linear Regression with Original Data Sets}
Figure \ref{02_GeneralRegression} shows the results of the linear regression analysis of each group separately, between the values of the observed stiffness tensors from \si{\micro}FE simulations and the predicted values using the Zysset-Curnier model\cite{Zysset1995} and the parameters obtained after performing the linear regression with linear mixed-effect model. The dashed line represents the 1:1 correlation. For the healthy group (Figure \ref{02_Healthy}) the fit is performed on 720 ROIs leading to 8640 data points. The $R^2_{adj}$ is slightly above 0.95 and the NE is of 18\% $\pm$ 10\%. The linear regression analysis for the OI group (Figure \ref{02_OI}) performed on 294 ROIs led to 3528 data points, an $R^2_{adj}$ close to 0.85 and a NE of 62\% $\pm$ 233\%. It can be noticed that, as the values of the observed stiffness tensor decreases, data points tend to be further apart from the diagonal (dashed line). Moreover, data the points from the stiffness tensors with lowest values are exclusively above the diagonal. The range of stiffness tensors of the OI group is wider compared to the one of the healthy group and ROIs with lower BV/TV present lower stiffness values. The values of these ROIs stiffness tensors components tend to be overestimated by the fit.\\

\begin{figure}[h!]
	\centering
	\begin{subfigure}[b]{0.5\textwidth}
		\centering
		\includegraphics[width=\textwidth]
		{Pictures/02_GR_Healthy_LMM}
		\caption{Healthy group}
		\label{02_Healthy}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.5\textwidth}
		\centering
		\includegraphics[width=\textwidth]
		{Pictures/02_GR_OI_LMM}
		\caption{OI group}
		\label{02_OI}
	\end{subfigure}
	\caption{Linear regression results using the fixed effects of the linear mixed-effect model on original data sets. $\lambda_{ii}$ stands for the diagonal terms of normal components of $\mathbb{S}$ in Mandel notation\cite{MANDEL1965}, $\lambda_{ij}$ for the off-diagonal terms of normal components, and $\mu_{ij}$ for the shear components. The dashed line represents the 1:1 correlation.}
	\label{02_GeneralRegression}
\end{figure}

\subsection{Filtering}
The CV in relation to BV/TV is shown in Figure \ref{02_CV_BVTV}. The OI samples reached higher CV values and lower BV/TV values compared to healthy ones. Generally, the CV tends to increase with decreasing BV/TV. The Spearman coefficient is shown above the plot as value [95\% CI]. Its value is negative and strictly different from zero. Finally, the CV threshold value used to filter the data is represented by the dashed line. It can be observed that a relatively important part of OI data will be filtered out. On the other hand, relatively few healthy data gets removed. 3D representation of extreme ROIs in terms of CV and BV/TV are shown in Appendix \ref{A2}.\\

\begin{figure}[h!]
	\centering
	\includegraphics[width=\linewidth]
	{Pictures/03_CV_BVTV}
	\caption{Coefficient of variation in relation to BV/TV. Spearman correlation coefficient $\rho$ assess monotonic relation between two variables}
	\label{02_CV_BVTV}
\end{figure}

The linear regression results of the filtered data are presented in Figure \ref{04_FilteredRegression}. After filtering, the healthy group was reduced to 119 individuals and 603 ROIs resulting in 7236 data points, an $R^2_{adj}$ close to 0.95 and a NE of 16\% $\pm$ 8\% (Figure \ref{04_Healthy}). In the OI group, more individuals were filtered, leading to 38 individuals and, 115 ROIs resulting in 1380 data points (Figure \ref{04_OI}), an $R^2_{adj}$ close to 0.95 and a NE of 17\% $\pm$ 8\%.\\

\begin{figure}[h!]
	\centering
	\begin{subfigure}[b]{0.5\textwidth}
		\centering
		\includegraphics[width=\textwidth]
		{Pictures/04_FR_Healthy_LMM}
		\caption{Healthy group}
		\label{04_Healthy}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.5\textwidth}
		\centering
		\includegraphics[width=\textwidth]
		{Pictures/04_FR_OI_LMM}
		\caption{OI group}
		\label{04_OI}
	\end{subfigure}
	\caption{Linear regression results using the fixed effects of the linear mixed-effect model on filtered data sets. $\lambda_{ii}$ stands for the diagonal terms of normal components of $\mathbb{S}$ in Mandel notation\cite{MANDEL1965}, $\lambda_{ij}$ for the off-diagonal terms of normal components, and $\mu_{ij}$ for the shear components. The dashed line represents the 1:1 correlation.}
	\label{04_FilteredRegression}
\end{figure}
